Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-08-12 document/ms.tex
GJH contributions from afternoon of Aug 12, some new figures added

\subsection{Rotational Modulation in the $V$-band lightcurve}\label{sec:rotSpot1}

The  $V$ brightness corresponds to a filling factor of the cool spot based on two temperatures fit from IGRINS spectra.  Although our analysis is consistent with a range of acceptable temperatures and filling factors, we adopt components of 2800 K covering 80\% of the visible stellar surface and 4200 K covering 20\% of the stellar surface at the time of the IGRINS spectrum, with an estimated $V_{rm est}=12.84$.  With these temperatures, the $V$-band emission is dominated by the hotter component.  The $V$-band brightness is therefore a good proxy for the visible surface area of the hot component.

Figure~\ref{fig:vband_spot} shows the recent ASAS-SN lightcurve, converted into a cool spot filling factor.  In these epochs, the brightest and coolest phase corresponds to cool component fill factors of $74\%$ and $86\%$, respectively.  Since 1992 the $V$ band photometry has been as bright as $12.3$ mag, corresponding to a cool component fill factor of $65\%$, and as faint as $13.2$ mag, or a cool component fill factor of $87\%$.  This drastic change in visible area of the hot spot (35\% to 13\%) is needed to explain the full $\sim 1$ mag range in brightness.

@@

Figures~\ref{fig:sed}-\ref{fig:lores} compares synthetic spectra from the two-component photosphere to the SED and flux-calibrated spectra of LkCa 4.
The only free parameters are the luminosity and extinction, which are both scaled to match the spectrum.
The optical-IR SED is obtained from estimating photometry from \citet{grankin08} during the 2MASS epoch.  We also included mid-IR photometry from Spitzer/IRAC \citep{hartmann05}, without adjusting for epoch.  In the spectroscopic comparison, some minor discrepancies may be introduced because the spot coverage changed in the $\sim 5$ hrs between the optical and near-IR spectra.  The synthetic spectrum is obtained from the Phoenix models, as in \S 3.

These comparisons require extinction estimates.  The $B-V$ color from \citet{grankin08} has an excess $E(B-V)=0.15$ mag relative to empirical colors of a 4200 K main sequence star and a $0.27$ mag excess for a pre-main sequence star \citep{kenyon95,pecaut13}.  The main sequence colors are probably most appropriate in this analysis because they are more free of spots.  Under this assumption, the $B-V$ color excess leads to an extinction $A_V=0.45\pm0.15$ mag, where the uncertainty range  corresponds to colors of 4100--4300 K temperature photospheres.   This extinction is overestimated if cooler temperatures also contribute to optical emission in a multi-temperature photosphere.  More accurate extinction estimates from the full SED are challenging because they depend on the areal distribution of temperatures in the photosphere, which in this work is simplied to two temperatures.  J--K discussion.  Bizarre V--R color note...

@@

Photometry is obtained from the main sequence colors compiled by \citet{kenyon95}.  Colors from dwarf stars should have minimal contribution from spots; measurements of pre-main sequence colors will include some unknown contribution of spots and cannot be used for this analysis.  Synthetic spectra of $3000-4000$ K stars predict $B-V$ colors that are much bluer than observed and that become bluer at cooler temperatures, the opposite of what is observed.  As a consequence, synthetic photometry also cannot be used here.

The optical emission is dominated by the hotter component while both components contribute to the red and infrared emission (see Figure \ref{fig:lores}).  The $0.7$ mag range in $V$-band brightness requires a factor of 1.9 change in the visible surface area of the hot component in the limit that the cool component does not contribute any $V$-band emission.  In the 2014-2015 ASAS-SN photometry, the minimum and maximum brightness correspond to cool spot coverage of 83\% and 68\%, respectively.  LkCa 4 was brighter in past epochs, which corresponds to either a larger covering fraction for the hot component (as assumed here) or components with hotter temperatures.

The rotational modulation leads to much smaller brightness changes at near-IR wavelengths than at optical wavelengths (see Table~\ref{tab:photrange}).  The smaller amplitude of near-IR brightness is also seen in a few spotted stars in the optical/IR monitoring of NGC 2264 \citep{cody14}, although those stars have a much smaller $V$ band amplitude than LkCa 4.  Figure~\ref{fig:colors} demonstrates that the $B-V$ color from \citet{grankin08} does not depend on $V$, while the $V-R_J$ becomes slightly redder when $V$ is smaller.  The predicted colors from rotational modulation do not match the observed $V-R_J$ colors or the correlation with $V$, either because of changes in spot properties since the \citet{grankin08} data were obtained or because a two-temperature model for the photosphere is overly simplistic.

\begin{figure}
 \centering
 \includegraphics[trim=3.1cm 13cm 2.3cm 3.6cm, clip=true, width=0.49\textwidth]{figures/phot1_paper}
\caption{The SED of LkCa 4 (red) compared with the two-temperature spectrum obtained from the best fit to the IGRINS spectra (black).}
\label{fig:sed}
\end{figure}

\begin{figure*}
 \centering
 \includegraphics[trim=2.1cm 13cm 2.5cm 3.1cm, clip=true, width=0.60\textwidth]{figures/lkca4_lores}
 \caption{The low-resolution optical/near-IR spectrum of LkCa 4 obtained from Palomar/DBSP and APO/Triplespec on 30 December 2008 (black), compared to a synthetic spectrum of a two temperature photosphere (purple).  The inset shows that the 3000 K (red, 70\% fill factor) and 4200 K (blue, 30\% fill factor) components contribute equally to the near-IR spectrum, but the 4200 K component dominates the blue emission.  The synthetic spectrum is reddened by $A_V=0.4$ mag and scaled to the observed $J$-band spectrum. \todo{Fix the $x-$axis label to be micron instead of Angstrom.} }
 \label{fig:lores}
\end{figure*}

\begin{figure*}
 \centering
 \includegraphics[trim=2.3cm 2.3cm 1.8cm 13.9cm, clip=true, width=0.75\textwidth]{figures/lkca4_lores_twopanel}
 \caption{The low-resolution optical (left) and near-IR (right) spectrum of LkCa 4, compared with a 3900 K photosphere (blue), a 3500 K photosphere, and the two temperature photosphere that best fit the IGRINS spectrum.  The synthetic spectra are scaled separately to the optical spectrum at 0.75 $\mu$m and to the near-IR spectrum at 1.5 $\mu$m.  Warm photospheres accurately reproduce molecular bands at $0.7$ $\mu$m but fail to fit the spectral features at longer wavelengths.  Cooler photospheres predict molecular bands at $<0.7$ $\mu$m that are much deeper than observed.  The two temperature photosphere accurately fits spectral features in the optical and near-IR.}
 \label{fig:lores_twopanel}
\end{figure*}

\begin{figure*}
 \centering
 \includegraphics[trim=1.6cm 12.6cm 2.2cm 2.4cm, clip=true, width=0.60\textwidth]{figures/vband_spot}
\caption{The $V$-band magnitude in 2014--2015, converted into fill factor for the cool component.  The optical brightness depends mostly on the hot component.  If we fix a 75\% filling factor, as measured in the IGRINS spectrum, to $V=12.83$ at the time of the observation, then the $V$-band amplitude  corresponds to filling factors of 67--83\%.  The factor of $\sim 2$ change in visible surface area of the hot component, from 33\% to 17\%, is required to produce the $\delta V=0.6$. \todo{Update with all available photometry or the derived multi-term fit} }
\label{fig:vband_spot}
\end{figure*}

@@

The depths of TiO and other molecular bands are common diagnostics of spectral type (and therefore temperature) in optical spectra \citep[e.g.][]{kirkpatrick91}.  In a 2-temperature photosphere, the depth of observed TiO bands will depend on the fractional coverage of the components.  In this subsection, we demonstrate that the TiO band depths vary with spot coverage.

\citet{donati14} spectroscopically monitored LkCa 4 with 12 spectra obtained over 14 days covering $\sim 4$ rotation periods.  The spectra obtained near the estimated maximum and minimum optical brightness are shown in the top panel of Figure~\ref{fig:tiovar}.  Small but significant changes are seen in the TiO bands.  The blue spectrum does not change.

The TiO band depths in these spectra are measured from spectral indices listed in Table~\ref{tab:tio}.  The TiO indices are also combined into a single index by  calculating the number of standard deviations each point is from the median value of the index, and then averaging the standard deviation.

@@

The multi-temperature photosphere may also resolve some of the long-standing problems related to discrepancies between observed and model colors of young stars.  Low mass young stars are redder than expected in the near-IR \citep[e.g.][]{tottle15}.  As a consequence, extinctions are high when measured by comparing near-IR colors to colors expected for optical spectral types.  For LkCa 4, the excess $E(J-K)$ color leads to an extinction of $1.04$ mag for a 4200 K photosphere and \citep{pecaut13} colors.  This extinction is consistent with the color and SED analysis of \citet{furlan06}.  However, this near-IR extinction is overestimated because the emission is generated from a composite spectrum and not only the hot component.  
 Our analysis in \S 4 suggests that $A_V\approx0.5$ mag, similar to the $A_V=0.69$ mag from \citet{kenyon95}.

These issues likely explain why near-IR extinctions are often higher than optical extinctions for young stars, and our results here generally support the lower extinctions obtained at optical wavelengths \citep[see analysis in][]{herczeg14}.  If the range in spot temperature/coverage in young stars is large, then extinctions measured from empirical pre-main sequence color scales may also be suspect.  The hottest temperature component dominates emission at short wavelengths and may provide the most accurate extinction estimate for non-accretors.  These color discrepancies are not problems with photospheric models or with the spectral type-effective temperature scale, although our results suggest that any single spectral type may be a misleading estimate of effective temperature.

This improved characterization of LkCa 4 should lead to more accurate estimates of mass and age from pre-main sequence tracks.  However, our effective temperature and luminosity yield a mass of 0.33 $M_\odot$ and an age of $\sim 0.5$ Myr in the \citet{baraffe15} evolutionary models.  This age is uncomfortably young and the mass is uncomfortably low.  LkCa 4 is not located deeply within a molecular cloud and is not associated with any nearby Class 0/1 stars, which would be expected for a 0.5 Myr star.  Dynamical measurements of masses of young stars with similar optical properties suggest that LkCa 4 should have a mass of $\sim 0.6-1.0$ M$_\odot$ \citep[e.g.][]{guilloteau14,czekala16,rizzuto16}.  Even though this is only a single data point, the inferred age and low mass may instead be evidence that strong magnetic fields are inhibiting convection in LkCa 4.  As a result, the surface is cooler and releases less radiation, slowing the contraction rate.  LkCa 4 then appears more luminous and cooler than expected from tracks that do not consider magnetic fields.   Inclusion of magnetic fields may help to resolve some of the age discrepancies between intermediate mass and low mass young stars \citep[e.g.][]{herczeg15}.

>>>>

\begin{figure*}
 \centering
 \includegraphics[trim=1.6cm 12.6cm 2.2cm 2.4cm, clip=true, width=0.60\textwidth]{figures/vband_spot_2800k}
\caption{The $V$-band magnitude in 2014--2015, converted into fill factor for the cool component.  The optical brightness depends mostly on the hot component.  If we fix a 75\% filling factor, as measured in the IGRINS spectrum, to $V=12.83$ at the time of the observation, then the $V$-band amplitude  corresponds to filling factors of 67--83\%.  The factor of $\sim 2$ change in visible surface area of the hot component, from 33\% to 17\%, is required to produce the $\delta V=0.6$. \todo{Update with all available photometry or the derived multi-term fit} }
\label{fig:vband_spot}
\end{figure*}

\subsection{The $V$-band lightcurve and spot coverage}\label{sec:rotSpot1}

The  $V$-band brightness corresponds to a filling factor of the cool spot based on two temperatures fit from IGRINS spectra.  We adopt photosphere components of 2800 K covering 80\% of the visible stellar surface and 4200 K covering 20\% of the stellar surface at the time of the IGRINS spectrum, with an estimated $V_{rm est}=12.84$.  These parameters are selected as roughly the median of the range of acceptable temperatures and filling factors from \S 3, and are selected blindly with respect to the ability to reproduce the broadband SED and spectra.

The $V$-band emission is dominated by the hotter component and is therefore a good proxy for the visible surface area of the hot component.  Figure~\ref{fig:vband_spot} shows the recent ASAS-SN lightcurve, converted into a cool spot filling factor.  In these epochs, the brightest and coolest phase corresponds to cool component fill factors of $74\%$ and $86\%$, respectively.  Since 1992 the $V$ band photometry has been as bright as $12.3$ mag, corresponding to a cool component fill factor of $65\%$, and as faint as $13.2$ mag, or a cool component fill factor of $87\%$.  This drastic change in visible area of the hot spot (35\% to 13\%) is needed to explain the full $\sim 1$ mag range in brightness.

\begin{figure}
 \centering
 \includegraphics[trim=3.1cm 13cm 1.0cm 3.6cm, clip=true, width=0.49\textwidth]{figures/phot1_paper}
\caption{The SED of LkCa 4 (red) compared with the two-temperature spectrum obtained from the best fit to the IGRINS spectra (black).}
\label{fig:sed}
\end{figure}

\begin{figure*}
 \centering
 \includegraphics[trim=2.1cm 3.0cm 1cm 7.7cm, clip=true, width=0.70\textwidth]{figures/lores_panels_paper}
 \caption{Top:  The low-resolution optical/near-IR spectrum of LkCa 4 obtained from Palomar/DBSP and APO/Triplespec on 30 December 2008 (black), compared to a synthetic spectrum of a two temperature photosphere (purple).  The inset shows that the 3000 K (red, 70\% fill factor) and 4200 K (blue, 30\% fill factor) components contribute equally to the near-IR spectrum, but the 4200 K component dominates the blue emission.  The synthetic spectrum is reddened by $A_V=0.4$ mag and scaled to the observed $J$-band spectrum. Bottom:  The low-resolution optical (left) and near-IR (right) spectrum of LkCa 4, compared with a 3900 K photosphere (blue), a 3500 K photosphere, and the two temperature photosphere that best fit the IGRINS spectrum.  The synthetic spectra are scaled separately to the optical spectrum at 0.75 $\mu$m and to the near-IR spectrum at 1.5 $\mu$m.  Warm photospheres accurately reproduce molecular bands at $0.7$ $\mu$m but fail to fit the spectral features at longer wavelengths.  Cooler photospheres predict molecular bands at $<0.7$ $\mu$m that are much deeper than observed.  The two temperature photosphere accurately fits spectral features in the optical and near-IR.}
 \label{fig:lores}
\end{figure*}

@@

Figures~\ref{fig:sed}-\ref{fig:lores} compares synthetic spectra from the two-component photosphere to the SED and flux-calibrated spectra of LkCa 4.  The only free parameters are the luminosity and extinction, which are both scaled to match the spectrum as described below.  The optical-IR SED is obtained from estimating photometry from \citet{grankin08} during the 2MASS epoch.  The SED also includes mid-IR photometry from Spitzer/IRAC \citep{hartmann05}, without adjusting for epoch.  In the spectroscopic comparison, some minor discrepancies between the optical and near-IR spectra may be introduced because the spot coverage changed in the $\sim 5$ hrs between observations.  The synthetic spectrum is obtained from the Phoenix models, as in \S 3, and is extended beyond the longest wavelength (5 $\mu$m) for calculating the bolometric luminosity.

These comparisons require extinction estimates.  The $B-V$ color from \citet{grankin08} has an 
excess $E(B-V)=0.15$ mag relative to empirical colors of a 4200 K main sequence star and a 
$0.27$ mag excess for a pre-main sequence star \citep{kenyon95,pecaut13}.  The $B-V$ color 
excess leads to an extinction $A_V=0.45\pm0.15$ mag, or $0.84\pm0.15$ mag relative to 
pre-main sequence colors.   The listed uncertainty corresponds to colors of 
4100--4300 K temperature photospheres.  The main sequence colors are probably most 
appropriate in this analysis because they are more free of spots.    
This extinction is overestimated if cooler temperatures also contribute to optical 
emission in a multi-temperature photosphere.  In the near-IR, the estimated $V=12.61$ 
at the time of the 2MASS observations leads to a template (main sequence) color 
$J-K_s=0.85$, with a range of $\sim 0.04$ for acceptable solutions with other 
temperatures and covering fractions. The observed $J-K=0.93$ leads to $A_V=0.47$ mag, 
with a range of 0.23 to account for the range of other photospheric solutions.  
The template near-IR color is more robust than optical colors to differences 
in the main sequence/pre-main sequence temperature scales because the pre-main 
sequence $J-K$ color is redder at 4200 K and bluer at 2800 K.  Baed on this 
analysis, an $A_V=0.5$ mag is adopted throughout this paper.  More accurate 
extinction estimates from the full SED are challenging because they depend on the 
areal distribution of temperatures in the photosphere, which in this work is 
simplified to two temperatures.  {\bf GJH note:  this last sentence and the 
generic approach tells the reader that we don't know how to do SED fitting, 
which sounds extremely embarrassing.  I think it's also correct.}

@@

The relative contributions from the hot and cool components change as LkCa 4 rotates, leading to rotational modulation of colors.  We model this rotational modulation using the main sequence colors compiled by \citet{pecaut13}, which should have less contribution from spots than their pre-main sequence colors.  Synthetic photometry obtained from Phoenix (or BT-Settl) spectra cannot be used here because the predicted $B-V$ colors are much bluer than observed.

The optical emission is dominated by the hotter component while both components contribute equally to the infrared emission (see Figure \ref{fig:lores}).  Figure~\ref{fig:colors} demonstrates that the $B-V$ color from \citet{grankin08} does not depend on $V$, which is consistent with expectations for a single hot component with no contributions from the cooler component.  The range of 0.03 mag in $B-V$ color indicates that the spot temperature is stable to 60 K.  The predicted colors from rotational modulation do not match the observed $V-R_J$ colors or the correlation with $V$, either because of changes in spot properties since the \citet{grankin08} data were obtained or because a two-temperature model for the photosphere is overly simplistic.

Rotational modulation is expected to lead to much smaller brightness changes at near-IR wavelengths than at optical wavelengths (see Table~\ref{tab:photrange}).  The smaller amplitude of near-IR brightness is also seen in a few spotted stars in the optical/IR monitoring of NGC 2264 \citep{cody14}, although those stars have a much smaller $V$ band amplitude than LkCa 4.

@@

The depths of TiO and other molecular bands are common diagnostics of spectral type (and therefore temperature) in optical spectra \citep[e.g.][]{kirkpatrick91}.  In a 2-temperature photosphere cool enough for molecules to form, the depth of observed TiO bands will depend on the fractional coverage of the components.  In this subsection, we demonstrate that the TiO band depths vary with spot coverage using 12 epochs of CFHT/ESPaDOnS spectra obtained over 14 days ($\sim 4$ rotation periods) by \citet{donati14}.

The spectra obtained near the estimated maximum and minimum optical brightness are shown in the top panel of Figure~\ref{fig:tiovar}.  Small but significant changes are seen in the TiO bands.  The blue spectrum does not change.  The TiO band depths in these spectra are measured from spectral indices listed in Table~\ref{tab:tio}  {\bf GJH:  Is this needed?}.  The TiO indices are also combined into a single TiO index by calculating the number of standard deviations each point is from the median value of the index and then averaging the standard deviation.

@@

{\bf GJH note:  I think this gets pretty rambling}

The multi-temperature photosphere may also resolve some of the long-standing problems related to discrepancies between observed and model colors of young stars.  Low mass young stars are redder than expected in the near-IR \citep[e.g.][]{tottle15}.  As a consequence, extinctions are high when measured by comparing near-IR colors to colors expected for optical spectral types.  For LkCa 4, the excess $E(J-K)$ color leads to an extinction of $1.04$ mag for a 4200 K photosphere and \citep{pecaut13} colors.  This extinction is consistent with the color and SED analysis of \citet{furlan06}.  However, this near-IR extinction is overestimated because the emission is generated from a composite spectrum and not only the hot component.  Our analysis in \S 4 suggests that $A_V\approx0.5$ mag, similar to the $A_V=0.69$ mag from optical photometry \citet{kenyon95} and $A_V=0.35$ mag from optical spectroscopy \citet{herczeg14}.

These issues likely explain why near-IR extinctions are often higher than optical extinctions for young stars \citep[see analysis in][]{herczeg14}.  Our results tentatively support the lower extinctions obtained at optical wavelengths, although the template colors used to calculate extinction is unclear.  In the \citet{pecaut13} conversions, late K spectral types lead to lower temperatures for pre-main sequence than main-sequence stars, and at the same temperature the B-V colors are redder.  The extinction to LkCa 4 and similar objects depends on whether these differences are caused by spots or gravity.

Uncertainty in colors may be solved with self-consistent comparisons to young stars with similar spectral types, if spot coverage is also similar.  However, if the range in spot temperature/coverage in young stars is large, then extinctions measured from empirical pre-main sequence color scales may also be suspect.  The hottest temperature component dominates emission at short wavelengths and may provide the most accurate extinction estimate for non-accretors.  These color discrepancies are not problems with photospheric models or with the spectral type-effective temperature scale, although our results suggest that any single spectral type may be a misleading estimate of effective temperature.

The improved characterization of the photospheric emission and radius of LkCa 4 should lead to more accurate estimates of mass and age from pre-main sequence tracks.  However, our effective temperature and luminosity yield a mass of 0.33 $M_\odot$ and an age of $\sim 0.5$ Myr in the \citet{baraffe15} evolutionary models.  This age is uncomfortably young and this mass is uncomfortably low.  LkCa 4 is not located deeply within a molecular cloud and is not associated with any nearby Class 0/1 stars, which would be expected for a 0.5 Myr star.  Dynamical measurements of masses of young stars with similar optical properties suggest that LkCa 4 should have a mass of $\sim 0.6-1.0$ M$_\odot$ \citep[e.g.][]{guilloteau14,czekala16,rizzuto16}.  Even though this is only a single data point, the inferred age and low mass may instead be evidence that strong magnetic fields are inhibiting convection in LkCa 4.  As a result, the surface is cooler and releases less radiation, slowing the contraction rate.  LkCa 4 then appears more luminous and cooler than expected from tracks that do not consider magnetic fields.   Inclusion of magnetic fields may help to resolve some of the age discrepancies between intermediate mass and low mass young stars \citep[e.g.][]{herczeg15}.
